---
title: "Programming with automatic differentiation"
subtitle: "NIMBLE 2023 Virtual Workshop"
author: "NIMBLE Development Team"
date: "January 2023"
output:
  slidy_presentation: default
  beamer_presentation: default
pandoc_args:
  colorlinks: true
  linkcolor: red
  urlcolor: red

---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
library(nimble)
#source(file.path("..", "examples", "DeerEcervi", "load_DeerEcervi.R"), 
#       chdir = TRUE)
eval <- TRUE
```

# Advanced uses of AD in nimble

- What variables and/or steps get "baked in" to AD tapes
- `reset` argument to `derivs` (aka `nimDerivs`)
- Holding variables out of AD tracking.
- Understanding `if`-`then`-`else` and `for`-loops in AD
- More about using AD with models
- Using AD in user-defined functions and distributions
- Generic parameter transformations
- Double-taping for potential efficiency

Note: For `a^b`, if `b` will always be an integer, use `pow_int(a, b)` instead of `pow(a,b)` (which is identical to `a^b`).  If `b` is a hard-coded integer, `pow_int` will be used automatically.

# When values are "baked in" to AD tapes

If a variable is not subject to derivative-tracking, its value will enter the AD tape during recording and can't be changed until the tape is reset.

Example: `y = exp(c * x^2)` where `c` will not be tracked for derivatives.

```{r, eval=eval}
nf1 <- nimbleFunction(
  # Setup argument c will become member data of a C++ class.
  # It will not have derivative-tracking.
  setup = function(c) {},
  methods = list(
    foo = function(x = double()) {
      return(exp(c * x^2))
      returnType(double())
    },
    derivs_foo = function(x=double()) {
      d <- derivs(foo(x), wrt = "x", order = 0:1)
      return(d)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = 'foo'
)
```

# When values are "baked in" to AD tapes (cont.)

```{r, eval=eval}
nf1obj <- nf1(0.1) # c = 0.1
Cnf1obj <- compileNimble(nf1obj)
# Compare compiled and uncompiled
nf1obj$derivs_foo(3.)
Cnf1obj$derivs_foo(3.)
# Always test with an input not used for recording the tape
nf1obj$derivs_foo(2.5)
Cnf1obj$derivs_foo(2.5)
# Change the value of c
nf1obj$c <- 0.2
Cnf1obj$c <- 0.2
# Compare compiled and uncompiled
nf1obj$derivs_foo(2.5)
Cnf1obj$derivs_foo(2.5) # WRONG!!!
```

The last `Cnf1obj$derivs_foo(2.5)` gives the **wrong answer** because the tape has `c = 0.1` baked in.

# How to reset tapes with new baked-in values

```{r, eval=eval}
nf2 <- nimbleFunction(
  setup = function(c) {},
  methods = list(
    foo = function(x = double()) {
      return(exp(c * x^2))
      returnType(double())
    },
    derivs_foo = function(x=double(),
                          reset = logical(0, default = FALSE)) {
      d <- derivs(foo(x), wrt = "x", order = 0:1, reset = reset)
      return(d)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = 'foo'
)
```

# How to reset tapes with new baked-in values (cont.)

```{r, eval=eval}
nf2obj <- nf2(0.1) # c = 0.1
Cnf2obj <- compileNimble(nf2obj)
# Compare compiled and uncompiled
nf2obj$derivs_foo(3.)
Cnf2obj$derivs_foo(3.)
# Test with an input not used for recording the tape
nf2obj$derivs_foo(2.5)
Cnf2obj$derivs_foo(2.5)
# Change the value of c
nf2obj$c <- 0.2
Cnf2obj$c <- 0.2
# Compare compiled and uncompiled
nf2obj$derivs_foo(3.)
Cnf2obj$derivs_foo(3., reset=TRUE) # Record a new tape
# Test with an input not used for recording the tape
nf2obj$derivs_foo(2.5)
Cnf2obj$derivs_foo(2.5) ## CORRECT
```

# Holding variables out of AD tracking 

We may want to include `c` as an argument but gain efficiency by never tracking derivatives through it if we don't need to.

We will illustrate two ways to do this, using $c = c_1 c_2$ so there are two constants to ignore for derivative tracking.

```{r, eval=eval}
nf3 <- nimbleFunction(
  setup = TRUE,
  methods = list(
    foo = function(x = double(),
                   c1 = double(),    # Below we say to ignore c1
                   c2 = integer()) { # Integers never have derivatives tracked
      return(exp(c1 * c2 * x^2))
      returnType(double())
    },
    derivs_foo = function(x=double(),
                          c1 = double(),
                          c2 = integer(),
                          reset = logical(0, default = FALSE)) {
      d <- derivs(foo(x, c1, c2), wrt = "x", order = 0:1, reset = reset)
      return(d)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = list(foo = list(ignore = "c1"))
)
```

# Holding variables out of AD tracking (cont)

```{r, eval=eval}
nf3obj <- nf3()
Cnf3obj <- compileNimble(nf3obj)
# Compare compiled and uncompiled
nf3obj$derivs_foo(3.,  0.05, 2)
Cnf3obj$derivs_foo(3., 0.05, 2)
# Test with an input not used for recording the tape
nf3obj$derivs_foo(2.5, 0.05, 2)
Cnf3obj$derivs_foo(2.5, 0.05, 2)
# Use new values of c1 and/or c2:
nf3obj$derivs_foo(2.5, 0.1, 2)
Cnf3obj$derivs_foo(2.5, 0.1, 2) # ERROR: old values of c1 and c2 are in the tape
Cnf3obj$derivs_foo(3., 0.1, 2, reset=TRUE) # Record a new tape (with any value of x)
Cnf3obj$derivs_foo(2.5, 0.1, 2) # CORRECT: with c1=.1 and c2 = 2
```

# Understanding `if`-`then`-`else` and `for`-loops in AD

Any conditional branching or looping is baked into the tape (aka "unrolled" in the tape).

Here is a strange way to calculate either $x^p$ or $(0.5 x)^p$.

```{r, eval=eval}
nf4 <- nimbleFunction(
  setup = TRUE,
  methods = list(
    foo = function(x = double(),
                   p = double(),
                   half = logical()) { # Will be an integer, but say I use double()
      ans <- 1
      if(!half) {
        for(i in 1:p) {
          ans <- ans*x
        }  
      } else {
        for(i in 1:p) {
          ans <- ans*(0.5*x)
        }
      }
      return(ans)
      returnType(double())
    },
    derivs_foo = function(x=double(),
                          p = double(),
                          half = logical(),
                          reset = logical(0, default = FALSE)) {
      d <- derivs(foo(x, p, half), wrt = "x", order = 0:1, reset = reset)
      return(d)
      returnType(ADNimbleList()) 
    }
  ),
  buildDerivs = list(foo = list(ignore = c('p','i'))) # Necessary to avoid an error
)
```

# Understanding `if`-`then`-`else` and `for`-loops in AD (cont.)

```{r, eval=eval}
nf4obj <- nf4()
Cnf4obj <- compileNimble(nf4obj)
# Compare compiled and uncompiled
nf4obj$derivs_foo(1.5, 3, FALSE)
Cnf4obj$derivs_foo(1.5, 3, FALSE)
# Test with an input not used for recording the tape
nf4obj$derivs_foo(2.5, 3, FALSE)
Cnf4obj$derivs_foo(2.5, 3, FALSE)
# Use new values of p and/or half:
nf4obj$derivs_foo(2.5, 4, TRUE)
Cnf4obj$derivs_foo(2.5, 4, TRUE) # ERROR: old values of p and half are used
Cnf4obj$derivs_foo(1.5, 4, TRUE, reset=TRUE) # Record a new tape (with any value of x)
Cnf4obj$derivs_foo(2.5, 4, TRUE) # CORRECT: with p=4 and half=TRUE
```

# More about using AD with models

- When recording a tape using a model, all values that will be used need to be copied (internally) from the regular model to the AD model.  

- You  need to provide information about which nodes are needed, in two categories: `constantNodes` (not the same as `constants` in `nimbleModel`) and `updateNodes`.

- The function `makeDerivsInfo()` can determine these for you for most cases.

- When using a tape recorded with a model, one can update values that might have changed (without resetting the tape).

Let's look at an example (simple model used earlier):

```{r, eval=FALSE}
model_code <- nimbleCode({
  y ~ dnorm(mean = mu, sd = sigma)
  mu <- intercept + slope * x
  intercept ~ dnorm(0, sd = 1000)
  sigma ~ dunif(0, 10)
})
```

- Say `y` is data.
- `model$calculate('y')` returns the log probability density of `y`.
- `model$calculate(c('mu','y'))` calculates `mu` and then returns the log probability density of `y`.
- Say we will want derivative(s) of `model$calculate('mu', 'y')` with respect to `slope`.
- Values needed: `y`, `intercept`, `slope`, `x`, and `sigma`.
- Values that might **not** change between derivative calls: `y`, `x`.  These are **constantNodes**.
- Values that might change between derivative calls: `intercept` and `sigma`. These are **updateNodes**.
- `slope` does not need to be in `updateNodes` because the "recorded" function will assign new value into the model.

# More about using AD with models (cont.)

Build the example model:

```{r, eval=eval}
model_code <- nimbleCode({
  y ~ dnorm(mean = mu, sd = sigma)
  mu <- intercept + slope * x
  intercept ~ dnorm(0, sd = 1000)
  sigma ~ dunif(0, 10)
})
m <- nimbleModel(model_code,
                 inits = list(intercept = 1.2, slope = 0.5,
                              sigma = 0.7),
                 data = list(y = 6, x = 8.3), # Notice I put x as data, not constants
                 buildDerivs = TRUE,
                 )
cm <- compileNimble(m)
```

Scalar constants are baked into a model and do not exist as nodes, so I made `x` be data for illustration.

# More about using AD with models (cont.)

```{r, eval=eval}
nf5 <- nimbleFunction(
  setup = function(model) {
    calc_nodes <- c('mu', 'y') # Normally, use model-generic programming, not hand-coded nodes
    constantNodes <- c('y','x')
    updateNodes <- c('intercept','sigma')
    param <- 'slope'
  },
  run = function(slope = double()) {
    temp <- numeric(value = slope, length = 1)
    values(model, param) <<- temp # c(slope) makes a scalar into a vector
    log_prob <- model$calculate(calc_nodes)
    return(log_prob)
    returnType(double())
  },
  methods = list(
    derivs_run = function(slope = double(),
                          reset = logical(0, default=FALSE)) {
      wrt <- numeric(length = 1, value = 1) # indices, not names
      ans <- derivs(run(slope), wrt = wrt,
                    order = 0:1,
                    updateNodes = updateNodes,
                    constantNodes = constantNodes,
                    model = model,
                    reset = reset)
      return(ans)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = 'run'
)
```

Notice that `updateNodes`, `model`, and `constantNodes` are needed for `derivs`, so `derivs_run` and `run` are tightly coupled functions. 

# More about using AD with models (cont.)

```{r, eval=eval}
nf5obj <- nf5(m)
Cnf5obj <- compileNimble(nf5obj, project = m)
# Note current values of mu in uncompiled and compiled models
m$mu
cm$mu
# Get derivatives
nf5obj$derivs_run(slope=0.7)
Cnf5obj$derivs_run(slope=0.7)
# Note that values of mu were updated
m$mu
cm$mu
# Get derivatives with new slope, after changing sigma
m$sigma <- 0.3
cm$sigma <- 0.3
nf5obj$derivs_run(slope=0.8)
Cnf5obj$derivs_run(slope=0.8)
# Note that values of mu were updated
m$mu
cm$mu
# Change a constantNode's value: DANGER!
m$y <- 9
cm$y <- 9
nf5obj$derivs_run(slope=0.8)
Cnf5obj$derivs_run(slope=0.8) # ERROR: old value of y is baked into tape
# Reset the tape to bake in the new values of constantNodes
Cnf5obj$derivs_run(slope=0.8, reset=TRUE) # CORRECT
```

# More about using AD with models (cont.)

This example illustrated `constantNodes` and `updateNodes` manually. In practice, `makeDerivsInfo` will do a good job of determining these.

```{r}
# Normally, use model-generic programming methods to determine calcNodes
makeDerivsInfo(m, wrtNodes = 'slope', calcNodes = c('mu','y'))
```

Here is how we would use this in a `nimbleFunction`.

```{r, eval=eval}
nf5b <- nimbleFunction(
  setup = function(model) {
    param <- 'slope'
    calc_nodes <- c('mu', 'y') # Normally, use model-generic programming, not hand-coded nodes
    # For a more general nimbleFunction,
    # make param and/or calc_nodes be arguments
    derivsInfo <- makeDerivsInfo(model, param, calc_nodes)
    constantNodes <- derivsInfo$constantNodes
    updateNodes <- derivsInfo$updateNodes
  },
  run = function(slope = double()) {
    temp <- numeric(value = slope, length = 1)
    values(model, param) <<- temp # c(slope) makes a scalar into a vector
    log_prob <- model$calculate(calc_nodes)
    return(log_prob)
    returnType(double())
  },
  methods = list(
    derivs_run = function(slope = double(),
                          reset = logical(0, default=FALSE)) {
      wrt <- numeric(length = 1, value = 1) # indices, not names
      ans <- derivs(run(slope), wrt = wrt,
                    order = 0:1,
                    updateNodes = updateNodes,
                    constantNodes = constantNodes,
                    model = model,
                    reset = reset)
      return(ans)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = 'run'
)
```

```{r, eval=eval}
nf5bobj <- nf5b(m)
# Run it once uncompiled to be sure there isn't a silly typo.
nf5bobj$derivs_run(0.8)
# This should work identically to nf5obj above.
```

# Using AD in user-defined functions and distributions

Simply add `buildDerivs=TRUE` to the `nimbleFunction` for the user-defined function or distribution.

(Or `buildDerivs = list(run = list(ignore = c('param1','param2')))` to leave `param1` and `param2` out of derivative tracking.)

# Double-taping of derivatives

We can record derivatives of the playing of an AD tape.

Let's consider `exp(x^4)` to be a bit more interesting.

```{r, eval=eval}
nf6 <- nimbleFunction(
  # Setup argument c will become member data of a C++ class.
  # It will not have derivative-tracking.
  setup = TRUE,
  methods = list(
    foo = function(x = double()) {
      return(exp(x^4))
      returnType(double())
    },
    gradient_foo = function(x=double(),
                            reset = logical(0, default=FALSE)) {
      d <- derivs(foo(x), wrt = 1, order = 1, reset = reset)
      g <- d$jacobian[1,]
      return(g)
      returnType(double(1))
    },
    derivs_gradient_foo = function(x = double(),
                                   order = integer(1),
                                   reset = logical(0, default=FALSE)) {
      d <- derivs(gradient_foo(x=x, reset=reset),
                  wrt = 1,
                  order = order,
                  reset = reset)
      return(d)
      returnType(ADNimbleList())
    }
  ),
  buildDerivs = c('foo', 'gradient_foo')
)
```

# Double-taping of derivatives (cont).

```{r, eval=eval}
nf6obj <- nf6()
cnf6obj <- compileNimble(nf6obj)
cnf6obj$gradient_foo(.8)
cnf6obj$derivs_gradient_foo(.8, order = 0:2)
```

# Double-taping of derivatives (cont).

When double-taping with `model$calculate` being taped, the `model`, `constantNodes` and `updatedNodes` need to be provided at the outer-most `derivs` call (and could be provided in the inner calls too).

# Generic parameter transformations

It is often useful to let algorithms operate on unconstrained parameter spaces. E.g. most optimization algorithms don't handle parameter boundaries well (or at all).  HMC would have problems with parameter boundaries.

In the AD beta-release, we provide a generic parameter transformation system.  It is motivated by AD use cases and supports derivative tracking, but it can be used for any purpose.

# Generic parameter transformations

Here is an silly toy model with some constrained parameters (and no data):

```{r, eval=eval}
toy <- nimbleModel(nimbleCode({
  a ~ dgamma(shape = 2, rate = 1)  # a >= 0
  b ~ dunif(10, 30)                # 10 <= b <= 20
  y ~ dnorm(b, sd = 5) 
  alpha[1:5] <- c(3, 6, 7, 6, 3)
  c[1:5] ~ ddirch( alpha[1:5] )    # sum(c) == 1
}), data = list(y = 15))
```

Say we want to optimize with respect to `(a, b, c[1:5])`.

The answer should be:

- a = 1
- b = 15
- c = (.1, .25, .3, .25, .1) (i.e. alpha / sum(alpha))

# Generic parameter transformations (cont.)

Let's see what the parameter transformation nimbleFunction does.

```{r, eval=eval}
nodes <- c('a','b','c[1:5]')
transform_toy <- parameterTransform(toy, nodes)
values <- c(3, 11, .2, .2, .2, .2, .2)
# Calculate transformed parameters that have no boundaries
transform_toy$transform(values)
```

What is the result?

- `log(a)`
- `logit((b-10)/(30-10))`
- `(logit(c[1]), logit(c[2]/(1-c[1])), logit(c[3]/(1-sum(c[1:2])), ...)`

```{r, eval=eval}
# Invert to original parameters
tr <- transform_toy$transform(values)
transform_toy$inverseTransform(tr)
```

# Generic parameter transformations (cont.)

Supported transformation cases:

1. scalar unconstrained
2. scalar semi-interval: (0, Inf)
3. scalar interval-constrained: (0, 1)
4. scalar semi-interval (-Inf, b) or (a, Inf)
5. scalar interval-constrained (a, b)
6. multivariate {normal, t}
7. multivariate {wishart, inverse-wishart}
8. multivariate dirichlet
9. LKJ (prior for covariance matrix, `dLKJ_corr_cholesky`)

# Example: Gradient-enabled optimization with transformations

```{r, eval=eval}
myMax <- nimbleFunction(
  setup = function(model, wrt) {
    calc_nodes <- model$getDependencies(wrt)
    derivsInfo <- makeDerivsInfo(model, wrt, calc_nodes)
    updateNodes <- derivsInfo$updateNodes
    constantNodes <- derivsInfo$constantNodes
    transformer <- parameterTransform(model, wrt)
  },
  # Use negative log Prob for convenience with optim
  run = function(p_trans = double(1)) {
    p <- transformer$inverseTransform(p_trans)
    values(model, wrt) <<- p
    log_prob <- model$calculate(calc_nodes)
    return(-log_prob)
    returnType(double())
  },
  methods = list(
    grad_run = function(p_trans = double(1)) {
      d <- derivs(run(p_trans), wrt = 1:length(p_trans), order = 1,
                  model = model, updateNodes = updateNodes,
                  constantNodes = constantNodes)
      return(d$jacobian[1,])
      returnType(double(1))
    },
    findMax = function(init_p = double(1)) {
      init_p_trans <- transformer$transform(init_p)
      optim_result <- optim(init_p_trans, fn = run, gr = grad_run,
                            method = "CG")
      opt_par <- transformer$inverseTransform(optim_result$par)
      optim_result$par <- opt_par
      return(optim_result)
      returnType(optimResultNimbleList())
    }
 ),
  buildDerivs = 'run'
)
```

# Example: Gradient-enabled optimization with transformations (cont)

Re-create the model with `buildDerivs=TRUE`.

```{r, eval=eval}
toy <- nimbleModel(nimbleCode({
  a ~ dgamma(shape = 2, rate = 1)  # a >= 0
  b ~ dunif(10, 30)                # 10 <= b <= 20
  y ~ dnorm(b, sd = 5) 
  alpha[1:5] <- c(3, 6, 7, 6, 3)
  c[1:5] ~ ddirch( alpha[1:5] )    # sum(c) == 1
}),
data = list(y = 15),
buildDerivs=TRUE)
```

# Example: Gradient-enabled optimization with transformations (cont)

Build and use the `myMax` `nimbleFunction`:

```{r, eval=eval}
myMaxObj <- myMax(toy, c('a','b','c[1:5]'))
compiled <- compileNimble(toy, myMaxObj)
set.seed(123)
compiled$toy$simulate()
init_p <- values(compiled$toy, c('a','b','c[1:5]'))
init_p
compiled$myMaxObj$findMax(init_p)
```

The result should have

- a = 1
- b = 15
- c = (.1, .25, .3, .25, .1)

(i-th mode of a Dirichlet is $\frac{\alpha_{i}-1}{\sum_{i} (\alpha_{i} - 1)}$.)
